Fused regression for network inference in multiple species

### 2015-08-30 paper1.tex
hi

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist enough data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine different data sets, lest we lose valuable information from previous studies. The problems of inferring networks in related species, processes, and using similar technologies likely are not independent ones, and benefit from the infusion of additional data.

@@

To learn gene regulatory networks, we employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple data sources of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each network separately. To take advantage of cross-dataset information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. Genes are not always regulated by the same transcription factors under different conditions. For example, a TF which promotes expression of a gene under a given process may repress that same gene at another point. Using data from these conditions to learn one network will prevent the identification of these contradictory regulatory interactions; it is therefore important to be able to combine data in a principled manner. We can use knowledge about the conservation of regulation to while leveraging th eadditional data to learn networks describing disparate processes. In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a transcription factor in one species is orthologous to a TF in a related species, and a gene in the first species is orthologous to a gene in the second, then the interaction between the TF and gene in the first species should be similar to the interaction between the orthologous TF, gene pair in the second species.

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. In this case, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\subsection{Fused gene regulatory networks}
We hypothesize that related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_S(\beta^A_{g,k} - \beta^B_{h,l})^2$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients and controls the tradeoff between fitting the expression data and producing a set of networks that conform to evolutionary constraints. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} \lambda_S(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})^2
\end{equation}

Where the second sum is over pairs of interactions with fusion constraints. Note that each term in the quantity being minimized is an L2 norm of a linear combination of gene expressions, so the entire problem can be solved through linear regression with a suitable design matrix. The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

@@

\subsection{Solving using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

@@

We can then define $W_{i,j} = \beta_{j \times N + i}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

@@

The fused multi-source regression problem can solved by constructing a multi-source design matrix $X$ and data matrix $Y$, and then vectorizing.
The regression problems associated with two columns $g,h$ of $Y$, where $g$ came from $Y_G$ and $h$ came from $Y_H$, are independent, unless there is a constraint chain of length $L$ with the form (where $x$ is arbitrary)

%$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, j_1, m_1, J_2, j_2, m_2) \rightarrow (J_2,j_2,m_2,J_3,j_3,_m3)...(J_P, j_P, m_P, H, j_{P+1}, h)$$
$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, x, m_1, J_2, j_2, m_2)$$

We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $G,i,j,$ and $-\lambda_S$ to the column associated with $J, i',j'$ for constraint $(G,i,j,J,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $b_{i,j} - b_{i',j,i}$.
These entries each contribute  $||b_{i,j} - b_{i',\{ji\}}||^{~2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

In the context of regression for GRN inference with multiple cell sources, each gene $j$ from source $1$ has constraints $\forall_{i \in \{TFs\}} (1, i, j, 2, i’, j’)$ where $i',j'$ are the same gene in a different cell line.
As a result, each gene has only one constraint and $L=2$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $L\sum_{i=0}^S K_i \times L \sum_{i=0}^S K_i$ matrix, taking $O((L\sum_{i=0}^S K_i)^3)$.
In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $p$ is the size of a column subset such that there is a constraint chain $L$ linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large $L$.

@@

\subsection{Alternative penalty functions}
Fusion constraints are intended to penalize dissimilarity between interactions thought to be evolutionarily analogous. Because fusion constraints are L2, interaction weights which differ from each other by a large amount are excessively penalized, which effectively ensures that fused interactions are assigned similar weights. This may be inappropriate for interactions which are identified based on orthology as being analogous, but which are no longer similar due to evolutionary changes. We propose that a saturating penalty may be useful for dealing with uncertainty about which interactions are conserved.

@@

\subsection{Optimization}
We use cross-validation to select optimal regularization parameters $\lambda_R$ and $\lambda_S$. In order to determine the relevant scale of the fusion weight $\lambda_S$, we can compute the correlation between pairs of coefficients linked by fusion constraints. The maximal value of $\lambda_S$ should be just large enough so that this correlation is near $1$.

>>>>

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist enough data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine different data sets, lest we lose valuable information from previous studies. The problems of inferring networks in related species, processes, and using different technologies likely are not independent ones, and benefit from the infusion of additional data.

The extent to which datasets are related is an important consideration when deciding to combine them. Abundant data in one species should be a useful source when learning the network of a related species. But often, how much and what of orthologous networks' structure is conserved, is not known a priori. We would like, then to be able to learn the conserved subnetwork in order to make better predictions using multi-source network inference.

@@

To learn gene regulatory networks, we employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple data sources of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each network separately. To take advantage of cross-dataset information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. Genes are not always regulated by the same transcription factors under different conditions. For example, a TF which promotes expression of a gene under a given process may repress that same gene at another point. Using data from these conditions to learn one network will prevent the identification of these contradictory regulatory interactions; it is therefore important to be able to combine data in a principled manner. We can use knowledge about the conservation of regulation while leveraging the additional data to learn networks describing disparate processes. In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a transcription factor in one species is orthologous to a TF in a related species, and a gene in the first species is orthologous to a gene in the second, then the interaction between the TF and gene in the first species should be similar to the interaction between the orthologous TF, gene pair in the second species. By utilizing information about conservation of gene function across data set, species, and biological process, we enable the usage of less directly relevant data and improve network inference.

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\subsection{Fused gene regulatory networks}
We hypothesize that related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $p_{\lambda_S}(\beta^A_{g,k} - \beta^B_{h,l})$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients and controls the tradeoff between fitting the expression data and producing a set of networks that conform to evolutionary constraints. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\end{equation}

where the second sum is over pairs of interactions with fusion constraints.

When the penalty $p_{\lambda_S}$ is the L2 norm, the entire problem can be solved through linear regression with a suitable design matrix. The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

@@

\subsection{Solving fused L2 problems using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

@@

We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

@@

The fused multi-source regression problem can solved by constructing a multi-source design matrix $X$ and data matrix $Y$, and then vectorizing.
The regression problems associated with two columns $g,h$ of $Y$, where $g$ came from $Y_1$ and $h$ came from $Y_2$, are independent, unless there is a constraint chain of length $L$ with the form (where $x$ is arbitrary)

%$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, j_1, m_1, J_2, j_2, m_2) \rightarrow (J_2,j_2,m_2,J_3,j_3,_m3)...(J_P, j_P, m_P, H, j_{P+1}, h)$$
$$(Y_1, j_1, g, Y_2, j_2, h) \rightarrow (Y_2, j_2, h, Y_3, j_3, k)$$

where the $j$s refer to transcription factors. In subproblem 1, $j_1$ regulates gene $g$; $j_1$ is orthologous to $j_2$, a transcription factor of subproblem 2, which regulates $h$, a gene which is orthologous to gene $g$. There exists a chain because ($j_1$, $g$) is a TF, gene pair which is fused with ($j_2$, $h$), which is fused to ($j_3$, $k$).
We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $Y_g,i,j,$ and $-\lambda_S$ to the column associated with $Y_h, i',j'$ for constraint $(Y_g,i,j,Y_h,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $b_{i,j} - b_{i',j'}$.
These entries each contribute  $||b_{i,j} - b_{i',j'}||^{~2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

In the context of regression for GRN inference with multiple cell sources, each gene $j$ from source $1$ has constraints $\forall_{i \in \{TFs\}} (1, i, j, 2, i’, j’)$ where $i',j'$ are the same gene in a different cell line.
As a result, each gene has only one constraint and the chain has length $L=2$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $L\sum_{i=0}^S K_i \times L \sum_{i=0}^S K_i$ matrix, taking $O((L\sum_{i=0}^S K_i)^3)$.
In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $p$ is the size of a column subset such that there is a constraint chain $L$ linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large $L$.

@@

\subsection{Adaptive fusion}
Fusion constraints are intended to penalize dissimilarity between interactions thought to be evolutionarily analogous. Because fusion constraints are L2, interaction weights which differ from each other by a large amount are excessively penalized, which effectively ensures that fused interactions are assigned similar weights. This may be inappropriate for interactions which are identified based on orthology as being analogous, but which are no longer similar due to evolutionary changes. We propose that a saturating penalty may be useful for dealing with uncertainty about which interactions are conserved.

@@

\subsection{Optimization}
We use glmnet to compute the regularization path for the
 ridge penalty. Given a range of $\lambda_R$ values, we fit models for each $\lambda_R$. We use cross validation to select the optimal parameter, the $\lambda_R$ value which minimizes the error of prediction on the leave out set.
